---
title: "ATAP Workshop: Folder Organisation"


output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: lumen
    code_folding: show
    highlight: pygments
---


## Introduction

Good folder organisation is one of the simplest but most important aspects of reproducible research. Structuring projects clearly makes it much easier to:

- navigate analyses
- locate outputs
- avoid overwriting files
- collaborate with others
- reproduce workflows later

Throughout this workshop we assume that analyses are being run inside an R Project. 

Start by creating a folder that can house all of this work along the lines of ATAP workshop. Open R and save a new project in this folder. 

Shooting from the hip in the beginning will come back to bite you as projects grow in complexity! 


## Check your R version

Before starting, it is useful to check the version of R that you are using.

```{r, results='hide'}
R.Version()
```

## Installing and Loading Packages

We will use the `pacman` package as a convenient way to install and load packages simultaneously. You can still use install.packages() and library, but `pacmam` does the heavy lifting if you don't have certain packages installed already.

```{r}
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")
}
```

Load the required package:

```{r}
pacman::p_load(purrr)
```

## Working Directories

A working directory is the folder where R will look for files and save outputs by default.

Check your current working directory:

```{r, results='hide'}
getwd()
```

Because we are working inside an R Project, the working directory is automatically set to the project root when the project is opened. So you can simple navigate all your folders once created with "data/detections.csv" for example

Hardcoding wd with setwd() limits portability of a project and becomes cumbersome when projects grow.

## Creating Project Folders

A clean project structure helps separate:
- raw data
- scripts
- plots
- tables
- manuscript files

We first create a path object pointing to the project location.

```{r}
folder_path <- "."
```

Because we are working inside an R Project, `"."` refers to the project root directory. Instead of hardcoding

## Define Folder Names

Choose the folders you want to create. Common examples include:
- `data`
- `scripts`
- `plots`
- `tables`
- `paper`

Remember:
- folder names are character strings and must be enclosed in quotation marks
- avoid accidental spaces in folder names

```{r}
folders <- c(
  "data",
  "scripts",
  "plots",
  "tables",
  "paper"
)
```

## Create the Folders

We can use `purrr::walk()` to iteratively create all folders in one call.

```{r}
purrr::walk(
  file.path(folder_path, folders),
  dir.create,
  recursive = TRUE,
  showWarnings = FALSE
)
```

## Check the Folder Structure

Finally, confirm that the folders were created successfully.

```{r}
list.dirs(folder_path, recursive = FALSE)
```

Easy and now we can get stuck into it with peace of mind that things are easy to find. 